---
title: "Making interactive web-apps in R using Shiny"
author: "Stuart Lacy"
date: "2022-03-02"
output: 
  binb::metropolis:
      keep_tex: false
classoption: "aspectratio=169"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

  - \alert{Allow others to explore the output of your R code through a web browser!}
  - \alert{Learning objective}: understand what Shiny is, the potential uses it has for research, and be able to create a basic app after this talk
  - Assumes you have some familiarity with R, but no experience with web development is required
  - \alert{Follow along} with the examples by cloning this repository within RStudio:
  - File -> New Project -> Version Control -> Git -> Repository URL: `https://github.com/stulacy/shiny-introduction.git`
  
## What is Shiny?

  - An R package for creating \alert{web-apps}
  - Developed by the RStudio/tidyverse team
  - Designed with a focus on \alert{data visualisation}

## Why would I want to use it?

  - Very \alert{quick development time} for initial proof of concepts
  - Provide means for non-R users to access your work
  - Don't need to know any HTML/CSS/JS, but can use those skills for \alert{customisation}
  - Clean and modern looking default appearance
  - Have access to R's extensive libraries
  - Simple \alert{hosting} options
  
## Possible use cases

  - Interactively visualize and \alert{explore} datasets (rather than generating PDFs with hundreds of figures)
  - \alert{Monitor} live data streams
  - Provide \alert{tools} to accompany published research (either methods, models, or data)
  - Establish an online presence
  - Provide visual interface to R code (i.e. automatic report generation)
  - Examples:
    - [COVID-19 dashboard](https://vac-lshtm.shinyapps.io/ncov_tracker/)
    - [Genome browser](https://gallery.shinyapps.io/genome_browser/)
    - [MRI Visualization](https://haozhu233.shinyapps.io/shinyMRI-contest/)
    - [Lego Mosaic Creator](https://rpodcast.shinyapps.io/shinylego/)

# Shiny code layout
  
## What does Shiny code look like?

\begin{itemize}
  \item<1-> Shiny apps are organised into \alert{2 files}:
  \item<1-> \alert{$ui.R$} (\textbf{U}ser \textbf{I}nterface)
  \begin{itemize}[<1->]
    \item Defines the appearance of the app to end users (\alert{frontend})
    \item Every UI is composed of multiple \alert{``widgets''}
    \item Widgets can be \alert{outputs} such as plots, tables, text
    \item And \alert{inputs} such as buttons, sliders, text fields, etc...
  \end{itemize}
  \item<2-> \alert{$server.R$}
  \begin{itemize}[<2->]
    \item Defines all \alert{backend} logic needed to process incoming inputs and generate the required outputs 
    \item Majority of your R code lives here
  \end{itemize}
  \item<3-> Input widgets are defined by \texttt{xInput} function calls in the UI
  \item<3-> Output widgets are referenced in both the UI (\texttt{xOutput} functions) and in the server (\texttt{renderX})
\end{itemize}
  
## Example 1: Basic working example

  - `examples/1_basic/`
  - Example application with a single plot and table
  - No interactivity at all
  - Can look at the code either on GitHub or can clone it to your PC [https://github.com/stulacy/shiny-introduction](https://github.com/stulacy/shiny-introduction)
  - Can run the app locally either by opening the project in RStudio, or can simply run `shiny::runGitHub(repo="shiny-introduction", username="stulacy", subdir="examples/1_basic")`

## Example 2: Interactive output elements
 
  - There are many options for \alert{interactive} output widgets:
    - `plotly`: general plotting library with controls
    - `DataTables`: interactive tables
    - `leaflet`: maps
    - `networkD3`: network diagrams
    - `diagrammeR`: graphs and flowcharts
    - `r2d3`: Interface to D3 Javascript plotting library
  - `examples/2_interactive_widgets`
  - The same layout as example 1, but with an interactive plot and table
  
# Reactivity

## Example 3: Reactive user inputs

  - The main form of adding interactivity in Shiny apps is through \alert{reactive} connections between the UI and Server
  - Input UI elements are referenced in `server.R` through the `input` object, e.g. the current value of a dropdown menu
  - Whenever this value changes due to user input, any output widgets that reference it will \alert{re-evaluate} and pass their new output (plot, table etc...) back to the UI
  - `examples/3_reactivity`

## Example 4: Multiple reactive dependencies

  - There is a \alert{many:many} relationship between input and outputs. Each `input` can be used in multiple outputs (as seen in previous example), and likewise each `output` can reference multiple `inputs`
  - There are lots of possible \alert{input widgets}:
    - Buttons
    - Checkboxes
    - Date selection/range
    - File upload
    - File download
    - Text input
    - Radio buttons
    - Dropdown menu
    - Sliders
  - `examples/4_multiple_reactives`
    
## Example 5: Custom reactive objects

  - So far we've seen that `input` objects are reactive and will \alert{trigger a change} in any downstream output that uses them
  - You can also \alert{create your own} reactive elements in a variety of ways
  - \alert{Be careful} when there are lots of reactive elements, it can make your code hard to follow and your app possibly slow!
  - `examples/5_custom_reactivity`
  
## Reactivity chaining

\begin{columns}[T,onlytextwidth]
    \column{0.5\textwidth}
      Example 4
      \begin{itemize}
        \item User selects transform or variable 
        \item Simultaneously:
        \begin{itemize}
          \item Scatter plot transforms selected variable and is redrawn
          \item Density plot transforms selected variable and is redrawn
        \end{itemize}
      \end{itemize}
    \column{0.5\textwidth}
      Example 5
        \begin{itemize}
            \item User selects transform or variable 
            \item $transformed\_df$ transforms selected variable
            \item Simultaneously:
            \begin{itemize}
              \item Scatter plot is redrawn
              \item Density plot is redrawn
            \end{itemize}
        \end{itemize}
\end{columns}

## More reactive objects

  - Alternatively can use the `observe` and `reactiveValues` paradigm, useful when have multiple reactive items or when the reactive object has stateful characteristics
  - This can be a complex topic, so refer to the [Shiny documentation](https://shiny.rstudio.com/articles/reactivity-overview.html) for more details

\footnotesize 

```{r eval=F, echo=T}
updated_vals <- reactiveValues()
observe({
  # Observe is triggered when any reactive expression contained within triggers
  # Unlike reactive, it doesn't return anything
  raw_val <- iris[[input$xvar]]
  funcs <- list("None"=identity, "Squared"=function(x) x**2, "Log"=log)
  
  transformed <- funcs[[input$transform]](raw_val)
  label <- sprintf("Displaying variable %s with transform %s", input$xvar, input$transform)
  updated_vals$data <- transformed
  updated_vals$label <- label
})
```
  

  
# Customising your app's appearance

## Structuring UI

  - By default all Shiny apps are \alert{responsive}, i.e. they adapt their layout to the size of the viewing device
  - To add more structure to an app, you can use a `sidebarLayout` to \alert{separate} inputs from outputs
  - You can partition your app further with \alert{tabbed output} using `tablistPanel`
  - You can create entirely independent \alert{pages} with `navbarPage`

## Example 6: UI Sidebar

  - `examples/6_ui_sidebar`
  - Shows an example of adding a \alert{sidebar} to an app
  - Useful for separating user controls from the resultant outputs
  
## Example 7: UI Tabs

  - `examples/7_ui_tabs`
  - Shows how separate UIs can be accessed through \alert{tabs}

## Example 8: UI Pages

  - `examples/8_ui_pages`
  - When a stronger distinction between different UIs is required, \alert{pages} can be used instead of tabs

## Further visual customisation

  - You can organise elements into rows (`fluidRow`) and columns (`column`) 
  - The \alert{bslib} package provides different themes if you get bored of the default
  - \alert{shinydashboard} package provides new UI elements for creating \alert{dashboard} style displays
  - Can add \alert{CSS} to have fine control over each element's appearance
  - Add flourishes with \alert{Javascript} (`shinyjs` package provides some useful features such as loading spinners)
  
## Example 9: Shinydashboard

  - `examples/9_shinydashboard`
  - Shows how the example 8 app looks under the `shinydashboard` package
  - The UI elements provided by `shinydashboard` are designed to get a quick dashboard up and running in a more structured manner than using the default UI

# Other considerations

## Organising larger apps

  - You can \alert{dynamically create UI elements} in the server end using `renderUI` and `uiOutput`
  - Code can start to get disorganised with larger multi-page apps - recommend putting each page into its own files and using `source` to load them in ([example here](https://github.com/stulacy/RDES-Shiny/blob/master/app.R)) or using [modules](https://shiny.rstudio.com/articles/modules.html)
  - If your program is running slower than expected, have a look at the \alert{reactive dependencies} and see if there is any redundancy
  - Bear in mind that each time a plot is updated through user interaction, a new plot needs to be generated and downloaded from the server - can be CPU intensive if done very frequently or a big dataset!

## Hosting: 

  - University provide \alert{free hosting} at shiny.york.ac.uk
  - 2 methods:
    - \alert{Managed}: provide access to a GitHub repo and it will automatically update whenever there is a change to the main branch
    - \alert{Self-hosted}: you are given access to a folder where you can put your app and any dependencies/data you need
  - Email `itsupport@york.ac.uk` for access
  - NB: not designed for a large number of concurrent users
  - Shinyapps.io has free hosting for up to 5 apps provided by RStudio team
  
## Data storage:

  - Several options for storing data
  - \alert{CSV/other structured files}
    - On same machine
    - Or remote (through AWS, Google Drive, or even Google Sheets)
  - \alert{Database} on same machine (SQLite) or remote (Postgres/MySQL/SQLServer etc...)
  - Databases are useful for big datasets as they only load the requested subsets into memory, rather than having to read the full file
  - Always worth doing as much pre-processing as possible \alert{offline}!

## Similar alternatives

  - Dash (Python)
  - Observable (Javascript/D3)
  - Tableau (drag and drop, commercial)
  - Grafana (drag and drop, open-source)
  
## Conclusions

  - Biggest \alert{strength}: Shiny makes it easy to get apps up and running with a wide range of features to cover most use cases
  - Biggest \alert{weakness}: For larger apps, the code can become hard to navigate, particularly if you start adding in custom JS/HTML/CSS
  - If you need support come along to a Research Coding Club drop-in session or ask on our slack channel `#research-coding-club`. Join the mailing list at [https://researchcodingclub.github.io/](https://researchcodingclub.github.io/)
  